---
title: "R1"
output: html_notebook
---

# Intro

Talking about `R` is kinda difficult for me. The main reason is that I don't precisely know what various CS terms mean. Moreover, I've become acquainted to the `R` philosophy which tries to do things, without regards for good practises and CS theory (and speed).

`R` has two meanings:
* a programming language specialised for data processing and statistical modelling
* the interpreter/shell which runs R commands (like IPython)

I suppose this doc will contain many misleading and incorrect uses of CS terms, so apologies for that.
Please, at any time, ask questions or say anything.

A little history. R is fairly modern (from 93') and is created as an open source ripoff of S, a Bell Labs product :)
From the basics, when it was imagined as a statistical programming language, it has grown into an extensive data-oriented system, providing many interesting stuff. Examples of those are:
* Shiny - R powered web app for working with data, modelling and visualising
* RMarkdown - reproducible notebook like documents
* plumber - R powered APIs
* ...

Also, there's been a lot of development of R which changed the way users write code and use R. We could say there's three schools of writing R:

* base R
* [tidyverse](https://www.tidyverse.org/)
* [data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)


### Usage

For running `R` interactively, I tend to use `Rstudio`. I suppose I started because it was similar to `MATLAB`, has a lot of `R`-specific things, debugging support, is project-oriented and so on.
I also use `VSC` for some interventions. It's now probably similar to RStudio in features, but I haven't though of switching.
For some short work, I sometimes run `R` directly from console.

For running `R` script in batch, use `Rscript`.


### Startup

The startup mechanism is a bit complex, but what you need to know for our repos is that `R` "sources" some specific files when started.
For instance, if you run `git clone git@github.com:optimoroute/biz-metrics.git; cd mrr` and start `R` - it will source `.Renviron` and then `.Rprofile`. The first is used for reading in environement variables, while the other contains `R` code to be run at each `R` startup.

To read more about this: `help(Startup)`


### Packages

R code comes bundles in packages. The official repository which includes some testing and checking is [CRAN](https://cran.r-project.org/).

CRAN packages are installed using `install.packages("ggplot2")`.
Packages can also be found on [Github](https://github.com/rstudio/renv) and you can install them using the `remotes` package: `remotes::install_github(rstudio/renv)`.

Loading (aka making the package namespace available) packages is done by `library(ggplot2)` or `require()`, but you can also use package objects by directly referencing them from the package namespace like `ggplot2::ggplot()`

By default, `R` saves downloaded packages to a directory obtained by `.libPaths()`.


###  Dependency management

I tend to use [renv](https://rstudio.github.io/renv/articles/renv.html) for managing packages. This works similar to `venv` and keeps a lockfile  called `renv.lock`. It's pretty new and still doesn't allow for managing different `R` versions.


### Getting help

You can obtain help from inside `R` on functions and many other topics using `?` or `help()`:

```{r}
help(max)

?`Startup`
```

CRAN based packages must have a website which contains a manual, see  e.g. [*data.table*](https://cran.r-project.org/web/packages/data.table/index.html)
There's a great thing called package vignettes. These are meant to be self-contained R notebooks
which explain how to use the package. See this intro vignette for [*data.table*](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)

*Tidyverse* packages tend to have [nice sites](https://ggplot2.tidyverse.org/)
with [great references](https://ggplot2.tidyverse.org/reference/index.html).


# R Landscape

### Basic operations

```{r}
# Assignment
z <- 10
z = 10 # convention: use this with function args

a <- b <- c <- 10
a

10 -> r -> e -> w 
w

# Arithmetics are all vectorised
# Recycling is done if necessary
1:10 + 1:10
1:10 + 1
1:10 + 100:101
1:10 * 11:20

sample(1:100, 10) >= 10

1:10 %% 2 == 0

# Logical operators - vectorised
1:10 %% 2 == 0 & 1:10 == 4
1:10 %% 2 == 0 | 1:10 == 3

# short circuiting (don't use this with vectors)
FALSE && {message("second"); TRUE}
FALSE & {message("second"); TRUE}

TRUE || {message("second"); TRUE}
TRUE | {message("second"); TRUE}

# if you need one result based on vectors, use this:
any(1:10 %% 2 == 0 & 1:10 == 4)
all(1:10 %% 2 == 0 | 1:10 == 3)

```

### Intro

> * Everything that exists is an object.
> * Everything that happens is a function call.
>
> — John Chambers

There's two kind of objects:

* Vectors
  * Atomic Vectors
    * Length
      * scalar
      * multiple length
    * Data Types
      * integer
      * double (numeric)
      * character
      * logical
      * *raw*
      * *complex*
  * Lists
* Nodes
  * Functions
  * Environments
  

### Vectors!

Vectors and `NULL` are de facto all that represents data in `R`. So, everything - scalars, lists, data frames - everything is a vector.

`NULL` is not a vector per se, but is often used a zero-length vector.

Vectors be of length 1 or >1. Length == 1 are called *scalars*.

Vector elements can be:

* all of same data type - ATOMIC vectors
* of different types - LISTS

Vectors can also have **attributes** - this is like a *named* list of metadata which serves different purposes.
E.g. the attribute `dimension` turns vectors into matrices and arrays and the attribute `class` runs the S3 "OO" system.


#### Atomic Vectors

```{r}

# Scalars
TRUE
FALSE
1.2323
1.213e9
1L # integer (Long)
"string"

# scalars are length 1 vectors
is.vector(TRUE)
length(TRUE)

# Generate by data type
invisible({
  numeric(length = 1000)
  double(1000)
  integer(1000)
  logical(1000)
  character(1000)
})

# Going longer
1:10 # same as 1:10L
c(TRUE, FALSE)  # concatenate
rep(TRUE, 10)
seq(1, 10, 0.1)
seq(from = as.Date("2020-01-01"), to = as.Date("2020-01-31"), by = "day")

# c flattens - always makes an atomic vector
c(c(TRUE, FALSE), c(TRUE, FALSE))  

# data type
typeof(TRUE)
typeof(23.4)
typeof(2L)


# Missing values
# Missing values are not NULL!!!!!
# This is completely different
NA # generic

# Each type has its own missing value
# this can sometimes cause errors because a few functions use type checking
NA_character_
NA_real_
NA_integer_

# most of the time, missing values propagate - be careful
sum(1, 23, NA)
sum(1, 23, NA, na.rm = TRUE)

# Checking for NA cannot be done with ==

x <- c(NA, 5)
x == NA # this is wrong
is.na(x)


# Testing types: is.something(x)
is.integer(10)
is.integer(10L)

# Coercing: as.something(x)
as.integer(123.2)
as.double(123L)

# Coercing order:
# character < double < integer < logical
x <- c(TRUE, TRUE, FALSE)
sum(x) # number of TRUE
mean(x) # proportion of TRUE

```


#### Lists

Vectors with elements of different types. Tehnically, each element is a reference for another object.

```{r}
l <- list(
  1:3, 
  "a", 
  c(TRUE, FALSE, TRUE), 
  c(2.3, 5.9)
)

typeof(l)
str(l)  # structure of object


is.list(1:5)
as.list(1:5)
unlist(as.list(1:5)) # this can have weird results
```

#### Data Frames / Tibbles

Special sort of list - each list element (aka column/variable) must be of same length.
This is the basic building block of doing most stuff in `R`. It's a shame that it's sometimes very slow, which we'll probably address during this session

The original `data.frame` sometimes acts funny and unexpected. The `tibble` is a (lightweight) fully compatible alternative with better properties used in the Tidyverse. I use tibble's almost exclusively.

```{r}
library(tibble)

# helper function
get_mrr <- function(resources, pricing_plan) {
  c("ST" = 19, "BUS" = 29, "ENT" = 39)[pricing_plan] * resources
}

set.seed(101)
n <- 50
tbl <- tibble(
  dates = sample(seq.Date(from = as.Date("2015-01-01"), to = as.Date("2020-12-31"), by = "day"), size = n),
  resources = sample(1:10, size = n, replace = TRUE),
  pricing_plan = sample(c("ST", "BUS", "ENT"), size = n, replace = TRUE),
  mrr = get_mrr(resources, pricing_plan)
)

tbl$dates[1]
mean(tbl$mrr)
```


#### Objects/Details

##### Names
Objects can have names (for each element). This is used often, especially in lists/data.frames:
```{r}
height <- c(borna = 174, marin = 192)
height["borna"] == height[1]

df <- tibble(height = sample(160:200, 100, replace = TRUE), weight = sample(50:110, 100, replace = TRUE), bmi = weight / (height/100)^2)

mean(df$bmi)
```

##### What's going on with the woRld

The [lobstr](https://www.tidyverse.org/blog/2018/12/lobstr/)  is great for taking a look at how `R` consumes memory, evaluates code and other stuff.
Also, because `R` is really bad and unintuitive with copying and using memory, `tracemem()` is a cool tool to see when and what is copied.


#### Metadata and S3

##### Attributes

All vectors can have metadata called **attributes**. Those are name-value pairs of metadata attached to any vector. You can see/set one of them with `attr(x, "some_attribute") / attr(x, "some_attribute) <-`, see them all with `attributes(x)` or set them all with structure:
```{r}
a <- 1:3
attr(a, "x") <- "abcdef"
attr(a, "x")

a <- structure(
  1:3, 
  x = "abcdef",
  y = 4:6
)
attributes(a) # list of attributes
str(a)
```

Most common attributes are:
* `names`
* `dim`   ... vector dimensions (matrices, arrays)
* `class` ... powers S3 OO system

```{r}
x <- 1:9 # vector 1D
dim(x) <- c(3,3)
x # matrix 3x3
# equivalent to matrix(1:9, nrow = 3, ncol = 3)

dim(x) <- c(3,3,1)
x # array 3x3x1
```


##### S3 system

Examples: factors, dates

```{r}
x <- sample(c("male", "female"), 100, replace = TRUE)

x
str(x) # this is a char vector
attributes(x)

fx <- factor(x, labels = c("male", "female"), ordered = FALSE)


fx # why does this get printed differently
str(fx)
attributes(fx)

print(x)
print(fx) # print.factor
methods(print)
print
print.factor
```


### Subsetting vector

#### Atomics

* indexing goes from 1, not from 0
* 6 ways to subset atomic vectors
* 3 subsetting operators

```{r}
x <- 101:110
names(x) <- letters[1:(length(x))]

# positive ints
x[c(1, 3, 4, 4)]
x[c(1.1, 2.9)] # xD

# negative ints
x[-1]
x[-c(1,2)]
x[-(1:99)]

# booleans (logicals) 
x[c(TRUE, FALSE)] # recycling
x[x %% 3 == 0]

# nothing
x[]

# zero
x[0]

# names
x[c("a","e")]

```

#### Lists

```{r}
l <- as.list(1:10)
names(l) <- letters[1:(length(l))]


l[1:2] # list

l[[1]] # list elements
l[["a"]]
l$a


# also for dataframes
tbl$dates
tbl[tbl$dates == "2018-12-24", c("mrr", "resources")] # still a tibble
tbl[tbl$dates == "2018-12-24", c("mrr")] # still a tibble
tbl_df <- as.data.frame(tbl)
tbl_df[tbl_df$dates == "2018-12-24", c("mrr", "resources")] # still a data.frame
tbl_df[tbl_df$dates == "2018-12-24", c("mrr")] # this is not a data.frame any more :/
```

#### Subsetting with assignment

You can also subset and assign like this:
```{r}
x[x%%2 == 0] <- x[ x%%2 == 0] + 4
x
```


### Functions

#### Basics

Components of each function:
* formals ... list of args
* body ... code 
* environment ... where the function is defined

```{r}
# Function definition
add_two_numbers <- function(a, b) {
  a + b
}

# Calling functions
add_two_numbers # print functions
add_two_numbers(10, 11)
add_two_numbers(a = 10, b = 11)
add_two_numbers(b = 11, 10)

args <- list(a = 10, b = 11)
do.call(add_two_numbers, args)
```


#### Other stuff

* **Primitive functions** are some functions which are written in C. These are located only in the `base` package, written by R-core.
```{r}
typeof(sum)
typeof(`[`)
# can't see their components
body(sum)
formals(sum)
environment(sum)
```

* Functions are **first-class** citizens in R - they act like any other object

* **Anonymous** functions:
```{r}
l <- list(1:1000)
map(l, function(x) x^2)
map(l, ~ .^2)
```


* **Lexical** scoping

```{r}
x <- 10
g01 <- function() {
  x <- 20
  x
}

g02 <- function() {
  x
}

g01()
x
g02()

```


* Partial-name matching (it's here so we raise attention not to use this)

```{r}
g03 <- function(value) {
  value^3
}

g03(v = 3)

```


* Lazy evaluation

```{r}
g041<- function(value, cond) {
  cond
  value^3
} 

g042 <- function(value, cond) {
  value^3
}

g041(value = 3, cond = stop("Stop this right now!"))
g042(value = 3, cond = stop("Stop this right now!"))

```

* Function Forms
  * prefix
  * infix
  * replacement
  * special
  * Borna remember: everything that happens is a function call
  
```{r}
`%+%` <- function(a, b) paste0(a, b)

# prefix
`%+%`("borna", "spalat")
# infix
"borna" %+% "spalat"
```


### Control Flow

```{r}

# if-else
if (runif(1) < 0.5) {
  print("Yes")
} else if (Sys.Date() == "2020-11-26") {
  print("It's today")
} else {
  print("No luck :)")
}

# vectorised ifelse
x <- sample(c(TRUE, FALSE), 100, replace = TRUE)
ifelse(x, 1, 0)

# vectorised ifelseif
y <- sample(c(1:7), 100, replace = TRUE)
days_in_week <- dplyr::case_when(
  y < 6 ~ "radni_dan",
  y == 6 ~ "subota",
  TRUE ~ "nedjelja"
)


# Loops
for (day in days_in_week) {
  message("The day is: ", day)
}

for (i in 1:length(days_in_week)) {
  message("The day is: ", days_in_week[i])
}
for (i in seq_along(days_in_week)) {
  message("The day is: ", days_in_week[i])
}

# while (isTRUE(something)) {
#   do this and that
# }


```


### I/O

```{r}
# RDS - R's binary format:
x <- rnorm(1000)
if (!dir.exists("data")) dir.create("data")
saveRDS(x, "data/sample_x.RDS")
x_original <- readRDS("data/sample_x.RDS")

# csv
write.csv(x, "data/sample_csv.csv", row.names = FALSE)
df <- read.csv("data/sample_csv.csv")
typeof(df) # this is not a vector

# modern alternative
# readr::write_csv(x, "data/sample_csv.csv") # crashes
readr::write_csv(data.frame(x = x), "data/sample_csv2.csv")
df <- readr::read_csv("data/sample_csv2.csv")


# Fast reading of large csvs
# data.table::fread
# https://jozef.io/r917-fread-comparisons/

# JSON
# jsonlite::fromJSON


```


### Environments

### Conditions

* errors
* warnings
* messages
